Localization properties, observable character and computational reliability of the linear response in open systems

### 2018-06-11 AnalyticStructure_main.tex

In this section we review the fundamental equations that govern the linear response regime of molecular-like systems in vacuum. Due to their intrinsic nature, this class of systems is 
characterized by a potential energy localized in a finite region around system's atoms and free boundary conditions proper of the open systems are adopted, so the wave function is in 
principle free to explore the whole space.

>>>>

Let us start our discussion by revisting the impact that the Isolated BC have on the 
solutions of the time-independent Schr\"odinger Equation.
We here present well-known concepts of elementary quantum mechanics, under a perspective that
will turn out to be very useful in the forthcoming considerations about linear-response quantities.

\subsection{Long range behavior of the eigenvalue solution of the Schroedinger equation}
\label{SEopenSystem}
Consider a (one-body) wavefunction $\ket{\psi}$ solution of the equation
$\op H\ket{\psi}=\eps\ket{\psi}$, where we split the Hamiltonian operator into the sum of
a kinetic $\op T$ and a potential $\op V \equiv \op H - \op T$ term.
In view of analysing the impact of the BC, it is instructive to transform 
the Schroedinger Equation in a \emph{scattering} problem, by writing
\be\lb{SEHelmholtz1}
\ket{\psi} = \op G_H(\eps)\ket{\op V\psi}\;,
\ee
where we have introduced the Green function of the Helmholtz operator $\op G_H(\eps) = (\eps-\op T)^{-1}$,
resolvent of the free-particle hamltonian $H_s \equiv \op T$.

When describing the Ground State of a molecule, we suppose that the potential operator, even if non-local,
is associated to real-space elements $\bra{\r} \op V \ket{\r'}$ that can be considered as nonzero only on a
bounded domain. In other terms, a molecule can be associated to a ``scatterer'' that has a \emph{finite} 
spatial extension; if a state $\ket{\psi}$ is localized, it will be the same for the ket $\ket{\op V \psi}$, though 
perhaps in a larger region.
For a computational treatment of the problem, this property is fundamental in the 
choice of the numerical treatment; we know that basis functions that are tailored to express asymptotically vanishing 
wavefunctions are better suited for the solution of the problem.

This fact is made apparent by the expression of 
the kernel of $\op G_H(\eps)$ in the coordinate representation. 
When dealing with Isolated BC, it is easy to see that the solution of Eq.~\eqref{SEHelmholtz1}
can be expressed by:
\be\lb{HelmholtzKernelDef1}
\bra{\r} \op G_H(\eps) \ket{\r'} = \frac{1}{4\pi} \begin{cases}
\frac{e^{-\alpha|\r - \r' |}}{|\r- \r' |} \;,\; & {\textrm for}\; \eps  < 0\; \\ 
\frac{e^{\ii \alpha |\r-\r' |}}{|\r-\r' |} \;, & {\textrm for} \; \eps \geq 0
\end{cases} \;,
\ee
where $\alpha = \sqrt{2|\eps|}$.

From the above expressions we can see that, \emph{for molecular systems},
the value $\epsilon=0$
represents a \emph{threshold}~\footnote{The threshold value in general should correspond to asymptotic value of the potential at infinity.} 
energy that separates two different classes of solutions.
Negative-energy solutions are \emph{localized}, i.e. they exhibit bound-state behaviour far away from the scatterer.
These are \emph{bound} states and, by means of their normalizability, they are associated to \emph{discrete}, well-defined energies.
The bound states can be of course labelled by quantum numbers that are defined by the properties of the scatterer, namely the
symmetries of the potential $\op V$.
In a computational discretization of the Schroedinger equation, such states can be in principle expressed with arbitrary precision, 
provided that the numerical treatment offers the adequate level of completeness. 
Being discrete and well-identified, the bound-state energies are a property of the molecule and can be associated to \emph{observable}
quantities. The values of their energies can be compared among different treatments and it is in principle possible to
provide reference results on them.

On the other hand, positive energy solutions behave differently. 
Even though the potential is localized, their positive energy makes them behave 
as spherical waves and its convolution with the source term  gives raise to a delocalized wavefunction $\psi(\r)$ 
in the whole accessible space.
As they cannot be normalized they belong to the essential spectrum of the hamiltonian operator and form a \emph{continuum} of states.
Thanks to the Lippman-Schwinger relation of Eq.~\eqref{SEHelmholtz1} they can be put in bijection with the eigenstates of the 
free-particle hamiltonian $\op H_s$. We may say that their quantum numbers are determined by the Laplacian operator, 
regardless of the particular features of the potential.

It is easy to see from these concepts that a computational treatment that is tailored for the discretization of bound states
will not be effective for continuum states, due to the different long-range asymptotic behaviour - which in turn is a clear consequence of the 
BC of the problem. In a numerical treatment, states with $\epsilon > 0$ will have a value of their energy that will \emph{always}
depend on the computational setup.

A direct numerical evidence of this problem has been discussed in \cite{boffi2016}. The authors of this paper analyze the effects of a finite basis set on the 
occupied and virtual orbitals of a molecular system and establish a clear correspondence between the orbital localization character, and the independence of its properties (\emph{e.g.} energy) 
with respect to the basis set. It is shown that (localized) orbitals with a bound-state behavior in the long range constitute a genuine discrete set and a finite basis is 
able to provide a reliable description of these states if the sampled domain contains their natural sizes. On the contrary, unbound (delocalized) orbitals form a continuum of states in the 
true system with open boundaries. For these class of objects the presence of a finite computational domain has a deeper impact since introduces a fictitious discretization of the energy levels. 
Furthermore, the energies of the unbound orbitals are not stable with respect to a change of the size of the computational domain and approach to zero as long as it is increased.

% In this section we review the fundamental equations that govern the linear response regime of molecular-like systems in vacuum. Due to their intrinsic nature, this class of systems is 
% characterized by a potential energy localized in a finite region around system's atoms and free boundary conditions proper of the open systems are adopted, so the wave function is in 
% principle free to explore the whole space.
